---
title: "Data Wrangling Exercises - Instructions"
author: "365: Intro to Bayesian Analysis, Fall 2023"
output: html_notebook
---


```{r setup, warning=FALSE}
library(tidyverse)
library(babynames)
library(gapminder)
library(readxl)
library(lubridate)
library(ggbeeswarm)
```

(Adapted from Steve Buyske, _Data Wrangling_)

<br>
<br>

## Intro

This segment is not specific to Bayesian analyses, but is instead about handling data in R. The functions are mostly from the `tidyverse` package. There are other ways to do the tasks that we are about to work through, but the tidyverse functions are particularly nice---they offer a consistent and easily read syntax.

We will be going through the material via a series of exercises. I **strongly** recommend trying each exercise before seeing what I have done. I've added lots of blank space to reduce the temptation to look ahead.


<br>
<br>

## Exercise 1

* Open the `Data Wrangling Exercises` project.
* Open a new R Notebook.
* Delete all of the text from "This is an ..." on and save it with a name of your choice.
* Create a code chunk with a name of `setup` and load the `tidyverse`,  `babynames`,  `gapminder`, `readxl`, `lubridate`, and ``ggbeeswarm` packages. To load a package, type `library(tidyverse)`, for example, on a separate line.


You should have code that looks like lines 8--15 above.

<br>
<br>

## Datasets, data frames, and tibbles

"Dataset" is a generic term for, well, a set of data. "Data frame" refers to a type of object in R, which is basically a big rectangle of data with rows corresponding to observations and columns corresponding to variables. The columns can have different types of data (character, number, TRUE/FALSE, and so on). A tibble is a data frame with a few extra features, but basically it prints more nicely. There is one difference that sometimes matters, though:
If you pull out a single column of a data frame, by default you will get a vector, and if you pull out a single element of a data frame, you will get a single string, number, TRUE or FALSE, and so on. 
If you do the same in a tibble, you will get another tibble.

If need be, you can convert a tibble to a data frame with `as.data.frame()` and a data frame to a tibble with `as_tibble()`.

```{r tibble and data frame example}
example_tbl <- tibble(x = 1:20, y = letters[1:20])
example_df <- as.data.frame(example_tbl)

example_tbl

example_tbl[, 1]
example_df[, 1]

example_tbl[1, 2]
example_df[1, 2]
```


<br>
<br>

## Babynames example

The `babynames` package has a data frame called `babynames`, which looks like

```{r baby example}
babynames
```

---

Here's the code to restrict to the name `Quinn`, restrict to 1980 and later, and to plot with separate colors for girls and boys.

```{r quinn example}
babynames %>%
  filter(name == "Quinn", year >= 1980) %>%
  ggplot(aes(x = year, y = n, colour = sex)) +
     geom_line() +
     ggtitle("Babies named Quinn")
```

<br>
<br>

## Exercise 2

* Try making a similar plot of "Taylor". 
* Add points to your plot by adding the code `+ geom_point()`


```{r}
babynames %>%
  filter(name == "Taylor", year >= 1980) %>%
  ggplot(aes(x = year, y = n, colour = sex)) +
     geom_line() + geom_point() +
     ggtitle("Babies named Taylor")
```

You should have something like

```{r exercise 2}
babynames %>%
  filter(name == "Taylor", year >= 1970) %>%
  ggplot(aes(x = year, y = n, colour = sex)) +
     geom_line() +
     geom_point() +
     ggtitle("Babies named Taylor")
```

<br>
<br>

## `group_by()` and `summarize()`
* We often want to summarize some aspect of a data frame by a grouping variable, or variables. 
  * For example, how often were different baby names given this century? 
* The two functions `group_by()` and `summarize()` will help you do so.
* E.g., 

```{r babynames example 1}
babynames %>% filter(year >= 2000) %>% 
  group_by(name) %>% 
  summarize(total = sum(n)) 
```


* Notice the addition of `arrange()` at the end, to sort the data, combined with `desc()` which puts a variable in descending order.

```{r babynames example 2}
babynames %>% 
  filter(year >= 2000) %>% 
  group_by(name) %>% 
  summarize(total = sum(n)) %>%
  arrange(desc(total))
```


* Notice the new explicit use of `head()`, with n = 5 to show the first 5 entries.

```{r babynames example 3}
babynames %>% filter(year >= 2000) %>% 
  group_by(name) %>% 
  summarize(total = sum(n)) %>%
  arrange(desc(total)) %>%
  head(n = 5)
```

<br>
<br>

## Exercise 3

What were the five most common baby names between 1901 and 1920?

```{r}
babynames %>% filter(year >= 1901 & year <= 1920) %>% 
  group_by(name) %>% 
  summarize(total = sum(n)) %>%
  arrange(desc(total)) %>%
  head(n = 5)
```
You should have something like

```{r exercise 3}
babynames %>% filter(year >= 1901, year <= 1920) %>% 
  group_by(name) %>% 
  summarize(total = sum(n)) %>%
  arrange(desc(total)) %>%
  head(n = 5)
```


<br>
<br>

## select()

* You can use `select()` to select only some of the variables in the data frame
```{r select example}
babynames %>% 
  select(name, year, sex, n) %>% 
  head(n = 3)
```


* You can also use `select()` to "unselect" by using a negative sign:
```{r select example 2}
babynames %>% 
  select(-prop) %>% 
  head(n = 3)
```

<br>
<br>

## Adding derived variables with `mutate()`

Often I want to create a new variable in a data frame by doing some calculation on the existing variables.
The function to do so is `mutate()`. 

For example, suppose I want to add a variable for the gdp in the `gapminder` data frame by multiplying `pop` and `gdpPercap`

```{r mutate example}
gapminder_with_gdp <-
  gapminder %>%
  mutate(gdp = pop * gdpPercap)

gapminder_with_gdp
```

<br>
<br>

## Exercise 4

* Create a new data frame `gapminder_extended` by adding a new variable that is the log base 10 of `gdpPercap`.
* `log10()` calculates the log base 10; `log()` is the natural log.


```{r}
#Create a new data frame 'gapminder_extend'
gapminder_extended <-
  gapminder %>%
  mutate(gdp = log10(gdpPercap))

gapminder_extended

```


You should have something like
```{r exercise 4}
gapminder_extended <-
  gapminder %>%
  mutate(lgdpPercap = log10(gdpPercap))

head(gapminder_extended, n = 3)
```

By the way, you can define as many new variables as you want; just separate them by commas inside `mutate()`. 

<br>
<br>

## `pivot_wider()` and `pivot_longer()`

* Sometimes we want the data to be "wide" and sometimes we want it to be "tall" or "long".
* E.g., the `babynames` data frame is long, with many rows and few columns:
```{r}
babynames %>% 
  filter(year >= 2000) %>%
  select(-prop) %>% 
  head(n = 3)
```

The long format is almost always what you want for plotting and for other analyses, but people often prefer to look at wide data.
You can use the `pivot_wider()` function to go from long to wide. The animation https://www.dropbox.com/s/a6o75zj443b2rv3/tidyr-longer-wider-modified.gif?dl=0 from [Mara Averick](https://twitter.com/dataandme/status/1175913657907253254) illustrates what `pivot_wider()` and its companion, `pivot_longer()`, do.


```{r wider example}
babynames_wide <- babynames %>%
  filter(year >= 2000) %>%
  select(-prop) %>%
  pivot_wider(names_from = "year", values_from = "n") 

babynames %>%
  head(n = 3)

babynames_wide %>%
  head(n = 3)
```


* You can go from tall to wide with `pivot_wider()`
* the `names_from` argument will give new column names
* the `values_from` argument will give the values in those columns
* all other variables determine the unique rows
  * you may need to dump variables using select

<br>
<br>

## Exercise 5

* Take the `gapminder` data frame and keep only the `country`, `year`, and `lifeExp` variables
* Turn the resulting data frame into a wide data frame
* Assign the result to an object name of your choice
* Show the first 3 rows


```{r}
#take the 'gapminder' data frame and keep only the 'country', 'year' and 'lifeExp'
gapminder_wide <- gapminder %>%
  select(country, year, lifeExp) %>%
  pivot_wider(names_from = "year", values_from = "lifeExp") 

gapminder %>%
  head(n = 3)

gapminder_wide %>%
  head(n = 3)
```












































You should have something like

```{r exercise 5}
gapminder_wide <- gapminder %>%
  select(country, year, lifeExp) %>%
  pivot_wider(names_from = "year", values_from = "lifeExp")

gapminder_wide %>%
  head(n = 3)
```


* Go from wide to long with `pivot_longer()`
* the `cols` argument speicifies the columns to convert to long format (try `?select_helpers` for some of the ways to specify the columns)
* the `names_to` argument gives the variable name for the variable created from the columns
* the `values_to` argument gives the variable name for the variable of the values from those columns

```{r longer example}
babynames_wide %>%
  head(n = 3)

babynames_wide %>% 
  pivot_longer(
    cols = `2000`:`2017`,  
    names_to = "year", 
    values_to = "count")
```


<br>
<br>

## Exercise 6

* Change the wide version of gapminder that you saved and convert it to a long version.

```{r}
gapminder_wide %>%
  head(n = 3)

gapminder_wide %>% 
  pivot_longer(
    cols = `1952`:`2007`,  
    names_to = "year", 
    values_to = "count")
```












































You should have something like

```{r exercise 6}
gapminder_wide %>% 
  pivot_longer(-country,  names_to = "year", values_to = "life_expectancy")
```


The vignette at https://tidyr.tidyverse.org/articles/pivot.html does a very nice job of showing simple examples of pivoting followed by more extended examples showing the use of most of the optional arguments

<br>
<br>

## Reading Excel files with `read_excel()`

So far we've either used data frames that are part of packages or we've read comma-separated values files using `read_csv()`. 
It's also quite common to have data in Excel files. The simplest way is with the `readxl` package, which you've already loaded.
The key function itself is called `readxl()`

```{r readxl example}
penguins <- read_excel("penguins.xlsx")

penguins
```
The data itself comes from the `palmerpenguins` package (intended as an alternative to the widely used iris dataset).

The defaults for `read_excel()` are to read from the first sheet, to assume that there are column headers, and to read all of the sheet. 
Here's an example with changes from the defaults.

```{r readxl example 2}
penguin_latin <- read_excel("penguins.xlsx", sheet = 2, skip = 3, n_max = 3)

penguin_latin
```


<br>
<br>

## Merging data frames using `full_join()`

We often want to merge different data frame. Here's an example

```{r full_join example}
full_join(penguins, penguin_latin, by = c("species" = "Common name"))
```

There are is also a `left_join()` function, which includes only observations in the first data frame;  a `right_join()` function that is includes only observations in the second data frame; and `inner_join()` that includes only observations that are in both datasets.
The most commonly used join is `left_join()`, which you use this whenever you are adding data from another dataset.
You can find more details at https://rafalab.github.io/dsbook/joining-tables.html 

<br>
<br>

## Exercise 7

* Merge the data frames `band_members` and `band_instruments2`.


```{r}
#You should call the data frame first 
band_members

band_instruments2

full_join(band_members,band_instruments2, c = ("name" = "artist"))
```












































You should have something like
```{r exercise 7}
band_members

band_instruments2

full_join(band_members, band_instruments2, by = c("name" = "artist"))
```

<br>
<br>

## Dates with the `lubridate` package

This is a small thing, but dates can be really annoying. This vector consists of strings all referring to the same date.

```{r lubridate example 1}
halloween <- c("10/31/20", "10/31/2020", "Oct 31, 2020", "October 31, 2020")
```

You can convert those to a common format

```{r lubridate example 2}
mdy(halloween)
```

All you need to do is specify the order of the month, day, and year; there are functions `dmy()`, `ymd()`, and so on.


<br>
<br>

## Other geoms


We will close out with a very quick look at some other ways of looking at data using the `ggplot()` function. You can find many others by searching Help for functions starting with `geom_` or doing web searches of ggplot2 and whatever kind of figure you are looking for.


### Boxplots

```{r boxplot example}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = continent, y = log10(gdpPercap), fill = continent)) +
  geom_boxplot()
```

### Density

```{r density example}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x  = log10(gdpPercap), fill = continent)) +
  geom_density(alpha = 0.75)
```
### Violin

```{r violin example}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = continent, y = log10(gdpPercap), fill = continent)) +
  geom_violin() 
```


### Beeswarm

```{r beeswarm example}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = continent, y = log10(gdpPercap), colour = continent)) +
  geom_beeswarm()
```

### Beeswarm and boxplot

```{r beeswarm and boxplot example}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = continent, y = log10(gdpPercap), colour = continent)) +
  geom_boxplot() +
  geom_beeswarm()

```

### Beeswarm and violin

```{r beeswarm and violin example}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = continent, y = log10(gdpPercap), colour = continent)) +
  geom_violin() +
  geom_beeswarm()

```


### Points and 2d density

```{r points and density2d example}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = log10(gdpPercap), y = lifeExp)) +
  geom_density2d_filled() +
  geom_point()
```
